Comparison of Centralized Manual and Automated Gating Methods
========================================================

* We compare OpenCyto and flowDensity against centralized manual gating for the SeraCare Lyoplate samples on five staining panels.  

```{r load_libraries,echo=FALSE}
suppressPackageStartupMessages({
  require(tools);
  require(data.table)
  require(reshape)
  require(knitr)
  require(lme4)
  require(car)
  require(ggplot2)
  require(languageR)
  require(Kmisc)
                               })
```

```{r add_javascript}
options(markdown.HTML.header = unlist(
  sapply(system.file('misc', c('vignette.css', 'datatables.txt'), package = 'knitr'), readLines)
  )
)
```

```{r load_data,echo=FALSE,cache=TRUE,cache.extra=md5sum("../../data/MergedTables.rda")}
load("../../data/MergedTables.rda")
```
 
## B-cell panel

```{r bcell_cleanup,echo=2}
  BCELL <- data.table(BCELL)
  summary(BCELL)
```

* We see that the manual method has more rows and there are some `NAs` in the data

```{r bcell_inspect}
  unique(BCELL[is.na(Proportion),list(Center,File,Method)])
  unique(BCELL[Proportion>1,list(Center,Population,Method)])
```

* The `NAs` come from Yale, and the file is not defined. This seems to be some missing data.
* There are "proportions" greater than 1 for a population that is NA as well. 
* We'll remove these and see if the rest is complete

```{r bcell_removeNA,echo=FALSE,message=FALSE,results='asis'}
  BCELL <- na.omit(BCELL)
  m<-melt(BCELL,id=c("Sample","Center","Population","Method"),measure="Proportion")
  kable(cast(m,Method~Population),format='html',table.attr='id="bcell_balance"')
```
<br>
* Okay, now things look nicely balanced. We check if the range of the data makes sense for proportions. 


`range(m$value)=[``r range(m$value)``]`

* That looks as expected. We're good to go.

<script type="text/javascript" charset="utf-8">
  $(document).ready(function() {
    $('#bcell_balance').dataTable();
	} );
</script>

* The last thing we need to do is annotation the samples with a technical replicate id.

```{r bcell_techrep_annotate}
  BCELL[,Replicate:=gl(nrow(.SD),1),list(Sample,Center,Population,Method)]
```


## Mixed effects model for the B-cell panel

We want to model variability between centers, between subjects, and contrast gating methods for each cell population.

### Raw data

```{r bcell_boxplot,fig.width=10,fig.height=5,fig.cap="Boxplots of log proportions for each center and cell population by subject and gating method.",cache=TRUE,dpi=300}
  df<-cast(BCELL,Sample+Center+Method~Population+Replicate,value="Proportion")
  BCELL <- BCELL[,lp:=logit(Proportion,adjust=0.00001)]
  BCELL <- BCELL[,logp:=log(Proportion)]
  pops <- levels(BCELL$Population)
  setkey(BCELL,Population)
  ggplot(BCELL[pops[c(3,5,8)]])+geom_boxplot(aes(y=Proportion,x=Center,fill=Method))+facet_grid(Population~Sample,scales="free")+theme(axis.text.x=element_text(angle=45,hjust=1))+ggtitle("Raw B-cell data")
```

How we'll model this is the following. We'll have fixed effects for gating methods, cell populations and their interactions. That is becausewe want to esimate the effec of each gating method on each population.

We fit a random intercept for Sample and Center as well as for each level of Population:Center and Population:Sample. The idea here is that cell population estimates will vary from center to center and from sample to sample, by more than just a fixed offset. 

We fit the reponse (proportions) on the logit scale.

```{r bcell_anova,cache=TRUE}
#Estimate fixed effects for population and method and their interaction
#Random effects for center and sample, as random intercept for each population:center and population:Sample
  mer<-lmer(lp~Population*Method+(1|Center/Population)+(1|Sample/Population),BCELL[Population!="Lymphocytes"],REML=FALSE,verbose=FALSE)
```

### Model fits and residuals


```{r bcell_summarize_fitted,echo=FALSE,fig.height=6,fig.width=12,dpi=200}
d <- data.frame(fit=fitted(mer),BCELL[Population!="Lymphocytes",],res=resid(mer),pg=BCELL[Population!="Lymphocytes",Method:Population])

  ggplot(d)+geom_boxplot(aes(x=pg,y=fit,fill=Method))+theme(axis.text.x=element_text(angle=90,hjust=1))+scale_y_continuous("fitted")+scale_x_discrete("Population:Gating Method")+scale_fill_discrete("Method")+ggtitle("Fitted Values (logit proportions) vs Method and Cell Population")+facet_wrap(~Sample,ncol=3)
```

```{r bcell_summarize_residuals,echo=FALSE,fig.height=5,fig.with=15,dpi=200,cache=FALSE}
  ggplot(d)+geom_boxplot(aes(x=pg,y=res,fill=Method))+theme(axis.text.x=element_text(angle=90,hjust=1))+scale_y_continuous("residuals")+scale_x_discrete("Population:Gating Method")+scale_fill_discrete("Method")+ggtitle("Residuals vs Method and Cell Population")+facet_wrap(~Sample,ncol=3)
  
   ggplot(d)+geom_point(aes(x=fit,y=res))+ggtitle("Residuals vs Fitted")

  ggplot(d)+geom_point(aes(x=Proportion,y=plogis(fit),col=Method))+ggtitle("Fitted vs Observed Proportions")+facet_grid(Center~Sample,scale="free")+geom_abline(aes(0,1),lty=2)+theme_bw()
```

### Bias

```{r bcell_bias,echo=FALSE,fig.width=5,fig.height=3,dpi=200}
  d<-data.frame(d,fixed=predict(mer,REform=0))

 d2<-cast(melt(d,id=c("Sample","Center","Population","Replicate","Method"),measure="fixed"),...~Method)

  ggplot(d2)+geom_point(aes(x=plogis(Manual),y=plogis(OpenCyto),shape="OpenCyto",color=Population),size=3)+geom_abline(aes(0,1),lty=2)+geom_point(aes(x=plogis(Manual),y=plogis(flowDensity),shape="flowDensity",color=Population),size=3)+theme_bw()+ggtitle(wrap("B-cell Panel Bias. Fitted Effects of Manual vs Automated Gates",width=35))+scale_y_continuous("Automated Gating Estimate")+scale_x_continuous("Manual Gating Estimate")

# Relative bias
  d3<-data.table(d)[,list(Method,diff.rel.pct=100*(.SD[,fixed]-.SD[Method=="Manual",fixed])/(.SD[Method=="Manual",fixed])),list(Sample,Center,Population,Replicate)]

ggplot(unique(d3[Method!="Manual"]))+geom_bar(aes(x=Population,y=diff.rel.pct,fill=Method),position="dodge",stat='identity')+coord_flip()+theme_bw()+scale_x_discrete("Cell Population")+scale_y_continuous("Percent Relative Difference")+ggtitle("B-cell Panel Bias as Percent Relative Difference")

# CV for plasmablasts
ggplot(BCELL[Population%in%c("Naive","Plasmablasts"),list(CV=var(log(Proportion))),list(Center,Method,Population)])+geom_boxplot(aes(y=CV,x=Method))+ggtitle("Distribution of Plasmablast CV across Subjects within Centers, by Gating Method")+theme_bw()+facet_wrap(~Population,scale="free")
```

### Variability

```{r bcell_variance_components,echo=FALSE,fig.width=4,fig.height=4,dpi=200,message=FALSE}
  #Extract Variance Components
 vc <- VarCorr(mer)
 foo<-capture.output({v <- as.numeric(print(vc,comp="Variance")[,3]);
 names(v) <- print(vc,comp="Variance")[,1]})
 v <- 100*v[order(v,decreasing=TRUE)]/sum(v)
 v <- melt(v)
  setnames(v,"VarComp")
  v<-data.frame(v,Component=rownames(v))
  v$Component <- ordered(v$Component,levels=v$Component)
    #Reorder the levels by descending magnitude
  ggplot(v,aes(x=Component,y=VarComp))+geom_bar(stat='identity')+ggtitle("B-cell Panel Variance Components")+scale_y_continuous("% of Variance Explained")+scale_x_discrete("Variance Component")+theme_bw()+theme(axis.text.x=element_text(angle=45,hjust=1))
  
```

## Summary of B-cell Panel

We note several things: 
*  First, OpenCyto is slightly biased for the Plasmablast cell population. IT tends to overestimate it compared to the centralized manual gates.
* Second, most of the variability is sample-to-sample biological variability, followed by residual within-sample variation, and then center-to-center variation. 
* The most variable populations are the plasmablasts and the IgD+ subsets.

## T-cell panel.

```{r Tcell,echo=FALSE}
  TCELLS <- data.table(TCELLS)
  summary(TCELLS)
  TCELLS <- na.omit(TCELLS)
```

* There are some `NAs` again.

```{r tcell_balance,results='asis'}
  m<-melt(TCELLS,id=c("Sample","Center","Population","Method"),measure="Proportion")
  kable(cast(m,Method~Population),format='html',table.attr='id="tcell_balance"')
```
<br>
For some reason there are more observations from flowDensity and Manual gating than OpenCyto.

```{r tcell_centers,results='asis'}
  kable(cast(m,Method~Center),format='html',table.attr='id="tcell_centers"')
```
* And we see the reason is that OpenCyto gating was not done on CIMR.
* We'll drop CIMR for the purposes of the analysis.

The range of the data looks okay.

`range(m$value)=[``r range(m$value)``]`


<script type="text/javascript" charset="utf-8">
  $(document).ready(function() {
    $('#tcell_balance').dataTable();
  } );
</script>
<script type="text/javascript" charset="utf-8">
  $(document).ready(function() {
    $('#tcell_center').dataTable();
  } );
</script>

* Annotate the technical replicates.

```{r tcell_techrep_annotate}
  TCELLS[,Replicate:=gl(nrow(.SD),1),list(Sample,Center,Population,Method)]
```

## Mixed Model for T-cells

```{r tcell_rawdata}
  TCELLS<-TCELLS[Center!="CIMR"] #drop CIMR
  df<-cast(TCELLS,Sample+Center+Method~Population+Replicate,value="Proportion")
  TCELLS <- TCELLS[,lp:=logit(Proportion,adjust=0.00001)]
  TCELLS <- TCELLS[,logp:=log(Proportion)]
  pops <- levels(TCELLS$Population)
  setkey(TCELLS,Population)
  ggplot(TCELLS[pops[c(3,5,12)]])+geom_boxplot(aes(y=Proportion,x=Center,fill=Method))+facet_grid(Population~Sample,scales="free")+theme(axis.text.x=element_text(angle=45,hjust=1))+ggtitle("Raw T-cell data")
```

We fit the mixed model to the T-cell panel.

```{r tcell_mixedmodel}
  mer<-lmer(lp~Population*Method+(1|Center/Population)+(1|Sample/Population),TCELLS[Population!="Lymphocytes"],REML=FALSE,verbose=FALSE)
```



```{r tcell_summarize_fitted,echo=FALSE,fig.height=6,fig.width=12,dpi=200}
d <- data.frame(fit=fitted(mer),TCELLS[Population!="Lymphocytes",],res=resid(mer),pg=TCELLS[Population!="Lymphocytes",Method:Population])

  ggplot(d)+geom_boxplot(aes(x=pg,y=fit,fill=Method))+theme(axis.text.x=element_text(angle=90,hjust=1))+scale_y_continuous("fitted")+scale_x_discrete("Population:Gating Method")+scale_fill_discrete("Method")+ggtitle("Fitted Values (logit proportions) vs Method and Cell Population")+facet_wrap(~Sample,ncol=3)
```

```{r tcell_summarize_residuals,echo=FALSE,fig.height=5,fig.with=15,dpi=200,cache=FALSE}
  ggplot(d)+geom_boxplot(aes(x=pg,y=res,fill=Method))+theme(axis.text.x=element_text(angle=90,hjust=1))+scale_y_continuous("residuals")+scale_x_discrete("Population:Gating Method")+scale_fill_discrete("Method")+ggtitle("Residuals vs Method and Cell Population")+facet_wrap(~Sample,ncol=3)
  
   ggplot(d)+geom_point(aes(x=fit,y=res))+ggtitle("Residuals vs Fitted")

  ggplot(d)+geom_point(aes(x=Proportion,y=plogis(fit),col=Method))+ggtitle("Fitted vs Observed Proportions")+facet_grid(Center~Sample,scale="free")+geom_abline(aes(0,1),lty=2)+theme_bw()
```

### Bias

```{r tcell_bias,echo=FALSE,fig.width=5,fig.height=3,dpi=200}
  d<-data.frame(d,fixed=predict(mer,REform=0))

 d2<-cast(melt(d,id=c("Sample","Center","Population","Replicate","Method"),measure="fixed"),...~Method)

  ggplot(d2)+geom_point(aes(x=plogis(Manual),y=plogis(OpenCyto),shape="OpenCyto",color=Population),size=3)+geom_abline(aes(0,1),lty=2)+geom_point(aes(x=plogis(Manual),y=plogis(flowDensity),shape="flowDensity",color=Population),size=3)+theme_bw()+ggtitle(wrap("T-cell Panel Bias. Fitted Effects of Manual vs Automated Gates",width=35))+scale_y_continuous("Automated Gating Estimate")+scale_x_continuous("Manual Gating Estimate")

# Relative bias
  d3<-data.table(d)[,list(Method,diff.rel.pct=100*(.SD[,fixed]-.SD[Method=="Manual",fixed])/(.SD[Method=="Manual",fixed])),list(Sample,Center,Population,Replicate)]

ggplot(unique(d3[Method!="Manual"]))+geom_bar(aes(x=Population,y=diff.rel.pct,fill=Method),position="dodge",stat='identity')+coord_flip()+theme_bw()+scale_x_discrete("Cell Population")+scale_y_continuous("Percent Relative Difference")+ggtitle("T-cell Panel Bias as Percent Relative Difference")

# CV for CD8, CD4 CM, CD4 Activated
ggplot(TCELLS[Population%in%c("CD8 Effector Memory","CD8 Central Memory","CD8 Naive"),list(CV=var(log(Proportion))),list(Center,Method,Population)])+geom_boxplot(aes(y=CV,x=Method))+ggtitle("Distribution of fo CV across Subjects within Centers, by Gating Method")+theme_bw()+facet_wrap(~Population)
```

### Variability

```{r tcell_variance_components,echo=FALSE,fig.width=4,fig.height=4,dpi=200,message=FALSE}
  #Extract Variance Components
 vc <- VarCorr(mer)
 foo<-capture.output({v <- as.numeric(print(vc,comp="Variance")[,3]);
 names(v) <- print(vc,comp="Variance")[,1]})
 v <- 100*v[order(v,decreasing=TRUE)]/sum(v)
 v <- melt(v)
  setnames(v,"VarComp")
  v<-data.frame(v,Component=rownames(v))
  v$Component<-ordered(v$Component,levels=v$Component)
  #Reorder the levels by descending magnitude
  ggplot(v,aes(x=Component,y=VarComp))+geom_bar(stat='identity')+ggtitle("T-cell Panel Variance Components")+scale_y_continuous("% of Variance Explained")+scale_x_discrete("Variance Component")+theme_bw()+theme(axis.text.x=element_text(angle=45,hjust=1))
```

## Summary of T-cell Panel

A couple of the CD8 populations exhibit some bias, but again, they are more consistent across subjects and centers than the manual gating.

#######################################
#######################################




## T-helper Panel

```{r Thelper,echo=FALSE}
  THELPER <- data.table(THELPER)
  summary(THELPER)
  THELPER <- na.omit(THELPER)
  THELPER<-THELPER[!Population%in%c("Lymphocytes","CD3","CD4","CD8")]
```

* There are some `NAs` again.

```{r thelper_balance,results='asis'}
  m<-melt(THELPER,id=c("Sample","Center","Population","Method"),measure="Proportion")
  kable(cast(m,Method~Population),format='html',table.attr='id="thelper_balance"')
```
<br>

```{r thelper_centers,results='asis'}
  kable(cast(m,Method~Center),format='html',table.attr='id="thelper_centers"')
  kable(cast(m,Method~Population),format='html',table.attr='id="thelper_populations"')

```
Things look balanced, and we have removed the Lymphocytes, CD8, CD4, and CD3 cells since they are not terminal populations.

The range of the data looks okay.

`range(m$value)=[``r range(m$value)``]`


<script type="text/javascript" charset="utf-8">
  $(document).ready(function() {
    $('#thelper_balance').dataTable();
  } );
</script>
<script type="text/javascript" charset="utf-8">
  $(document).ready(function() {
    $('#thelper_center').dataTable();
  } );
</script>
<script type="text/javascript" charset="utf-8">
  $(document).ready(function() {
    $('#thelper_populations').dataTable();
  } );
</script>

* Annotate the technical replicates.

```{r thelper_techrep_annotate}
  THELPER[,Replicate:=gl(nrow(.SD),1),list(Sample,Center,Population,Method)]
```

## Mixed Model for T-helper panel

```{r thelper_rawdata}
  df<-cast(THELPER,Sample+Center+Method~Population+Replicate,value="Proportion")
  THELPER <- THELPER[,lp:=logit(Proportion,adjust=0.00001)]
  THELPER <- THELPER[,logp:=log(Proportion)]
  pops <- levels((THELPER$Population))
  setkey(THELPER,Population)
  ggplot(THELPER[pops[c(10,11,12)]])+geom_boxplot(aes(y=Proportion,x=Center,fill=Method))+facet_grid(Population~Sample,scales="free")+theme(axis.text.x=element_text(angle=45,hjust=1))+ggtitle("Raw T-helper data")
```

We fit the mixed model to the T-cell panel.

```{r thelper_mixedmodel}
  mer<-lmer(lp~Population*Method+(1|Center/Population)+(1|Sample/Population),THELPER[Population!="Lymphocytes"],REML=FALSE,verbose=FALSE)
```



```{r thelper_summarize_fitted,echo=FALSE,fig.height=6,fig.width=12,dpi=200}
d <- data.frame(fit=fitted(mer),THELPER[Population!="Lymphocytes",],res=resid(mer),pg=THELPER[Population!="Lymphocytes",Method:Population])

  ggplot(d)+geom_boxplot(aes(x=pg,y=fit,fill=Method))+theme(axis.text.x=element_text(angle=90,hjust=1))+scale_y_continuous("fitted")+scale_x_discrete("Population:Gating Method")+scale_fill_discrete("Method")+ggtitle("Fitted Values (logit proportions) vs Method and Cell Population")+facet_wrap(~Sample,ncol=3)
```

```{r thelper_summarize_residuals,echo=FALSE,fig.height=5,fig.with=15,dpi=200,cache=FALSE}
  ggplot(d)+geom_boxplot(aes(x=pg,y=res,fill=Method))+theme(axis.text.x=element_text(angle=90,hjust=1))+scale_y_continuous("residuals")+scale_x_discrete("Population:Gating Method")+scale_fill_discrete("Method")+ggtitle("Residuals vs Method and Cell Population")+facet_wrap(~Sample,ncol=3)
  
   ggplot(d)+geom_point(aes(x=fit,y=res))+ggtitle("Residuals vs Fitted")

  ggplot(d)+geom_point(aes(x=Proportion,y=plogis(fit),col=Method))+ggtitle("Fitted vs Observed Proportions")+facet_grid(Center~Sample,scale="free")+geom_abline(aes(0,1),lty=2)+theme_bw()
```

### Bias

```{r thelper_bias,echo=FALSE,fig.width=5,fig.height=3,dpi=200}
  d<-data.frame(d,fixed=predict(mer,REform=0))

 d2<-cast(melt(d,id=c("Sample","Center","Population","Replicate","Method"),measure="fixed"),...~Method)

  ggplot(d2)+geom_point(aes(x=plogis(Manual),y=plogis(OpenCyto),shape="OpenCyto",color=Population),size=3)+geom_abline(aes(0,1),lty=2)+geom_point(aes(x=plogis(Manual),y=plogis(flowDensity),shape="flowDensity",color=Population),size=3)+theme_bw()+ggtitle(wrap("T-helper Panel Bias. Fitted Effects of Manual vs Automated Gates",width=35))+scale_y_continuous("Automated Gating Estimate")+scale_x_continuous("Manual Gating Estimate")

# Relative bias
  d3<-data.table(d)[,list(Method,diff.rel.pct=100*(.SD[,fixed]-.SD[Method=="Manual",fixed])/(.SD[Method=="Manual",fixed])),list(Sample,Center,Population,Replicate)]

ggplot(unique(d3[Method!="Manual"]))+geom_bar(aes(x=Population,y=diff.rel.pct,fill=Method),position="dodge",stat='identity')+coord_flip()+theme_bw()+scale_x_discrete("Cell Population")+scale_y_continuous("Percent Relative Difference")+ggtitle("T-helper Panel Bias as Percent Relative Difference")

# CV for CD8, CD4 CM, CD4 Activated
ggplot(THELPER[Population%in%c("CD8 Th2","CD8 Th1","CD4 Th2","CD4 Th1"),list(CV=var(log(Proportion))),list(Center,Method,Population)])+geom_boxplot(aes(y=CV,x=Method))+ggtitle("Distribution of CV across Subjects within Centers, by Gating Method")+theme_bw()+facet_wrap(~Population)
```

### Variability

```{r thelper_variance_components,echo=FALSE,fig.width=4,fig.height=4,dpi=200,message=FALSE}
  #Extract Variance Components
 vc <- VarCorr(mer)
 foo<-capture.output({v <- as.numeric(print(vc,comp="Variance")[,3]);
 names(v) <- print(vc,comp="Variance")[,1]})
 v <- 100*v[order(v,decreasing=TRUE)]/sum(v)
 v <- melt(v)
  setnames(v,"VarComp")
  v<-data.frame(v,Component=rownames(v))
  v$Component<-ordered(v$Component,levels=v$Component)
  #Reorder the levels by descending magnitude
  ggplot(v,aes(x=Component,y=VarComp))+geom_bar(stat='identity')+ggtitle("T-helper Panel Variance Components")+scale_y_continuous("% of Variance Explained")+scale_x_discrete("Variance Component")+theme_bw()+theme(axis.text.x=element_text(angle=45,hjust=1))
```

## Summary of T-cell Panel

A couple of the CD8 populations exhibit some bias, but again, they are more consistent across subjects and centers than the manual gating.



#######################################
#######################################




## DC / Mono / NK Panel



#######################################
#######################################

## Treg Panel
